\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-15 chapters/bci.tex
Proofread mathematical notation

\chapter{Brain-computer interfaces}
\label{sec:bci}

@@

There is one caveat with these linear methods, however:
the number of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
is usually relatively high compared to the training dataset, since calibration time
should be kept minimal.
Therefore, strong regularization should be applied, such as covariance/scatter
shrinkage.
Regularization can also be performed by retaining structure in the data, such as in the
case for the multilinear methods presented by~\textcite{Lotte2018}.

@@

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity in the background noise to regularize the problem.
It does this by imposing a block-Toeplitz structure to the between-class
scatter $\mat{S}_\text{b}$ matrix of \ac{lda}.
This means
\begin{equation}
  \mat{S}_\text{w} = \begin{bmatrix}
    \mat{S}_\text{w}^{(1,1)} & \mat{S}_\text{w}^{(1,2)} & \cdots & \mat{S}_\text{w}^{(1,C)}\\
    \mat{S}_\text{w}^{(2,1)} & \mat{S}_\text{w}^{(2,2)} & \cdots & \mat{S}_\text{w}^{(2,C)}\\
    \vdots             & \vdots             & \ddots & \vdots            \\
    \mat{S}_\text{w}^{(C,1)} & \mat{S}_\text{w}^{(C,2)} & \cdots & \mat{S}_\text{w}^{(C,C)}\\
  \end{bmatrix}
\end{equation}
consists of $C\times C$ temporal scatter matrices
$\mat{S}_\text{w}^{(c_1,c_2)}\in\mathbb{R}^{S\times S}$
corresponding to a pair of channels $(c_1,c_2)$, such
that every temporal scatter matrix has the Toeplitz structure
\begin{equation}
  \mat{S}_\text{w}^{(c_1,c_2)} = \begin{bmatrix}
    s_1^{(c_1,c_2)}  & s_2^{(c_1,c_2)}  & \cdots & s_S^{(c_1,c_2)}  \\
    s_2^{(c_1,c_2)}  & s_1^{(c_1,c_2)}  & \cdots & s_{S-1}^{(c_1,c_2)}  \\
    \vdots & \vdots &  \ddots & \vdots \\
    s_S^{(c_1,c_2)}  & s_{S-1}^{(c_1,c_2)}  & \cdots & s_1^{(c_1,c_2)}
\end{bmatrix}
\end{equation}
The structuring is performed by setting all elements to the average of their
corresponding block-subdiagonal and subsequently applying shrinkage
regularization.
This method is simple yet effective, reaching state-of-the-art decoding performance.

>>>>

\chapter{Brain-computer interfaces}
\label{sec:bci}
\epigraph{``My mind controls my reality.''}{Frank Herbert -- Children of Dune}

@@

There is one caveat with these linear methods, however:
the number of values ($=C\cdot S$) is usually relatively high compared to the
size of the training dataset, since calibration time should be kept minimal.
Therefore, strong regularization, such as covariance/scatter
shrinkage, should be applied.
Regularization can also be performed by retaining structure in the data, such as in the
case for the multilinear methods presented by~\textcite{Lotte2018}.

@@

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity in the background noise to regularize the problem.
It does this by imposing a block-Toeplitz structure to the within-class
scatter $\mat{S}_\text{w}$ matrix of \ac{lda}.
This means
\begin{equation}
  \mat{S}_\text{w} = \begin{bmatrix}
    \mat{S}_\text{w}^{(1,1)} & \mat{S}_\text{w}^{(1,2)} & \cdots & \mat{S}_\text{w}^{(1,C)}\\
    \mat{S}_\text{w}^{(2,1)} & \mat{S}_\text{w}^{(2,2)} & \cdots & \mat{S}_\text{w}^{(2,C)}\\
    \vdots             & \vdots             & \ddots & \vdots            \\
    \mat{S}_\text{w}^{(C,1)} & \mat{S}_\text{w}^{(C,2)} & \cdots & \mat{S}_\text{w}^{(C,C)}\\
  \end{bmatrix}
\end{equation}
consists of $C\times C$ temporal scatter matrices
$\mat{S}_\text{w}^{(c_1,c_2)}\in\mathbb{R}^{S\times S}$
corresponding to a pair of channels $(c_1,c_2)$, such
that every temporal scatter matrix has the Toeplitz structure
\begin{equation}
  \mat{S}_\text{w}^{(c_1,c_2)} = \begin{bmatrix}
    s_1^{(c_1,c_2)}  & s_2^{(c_1,c_2)}  & \cdots & s_S^{(c_1,c_2)}  \\
    s_2^{(c_1,c_2)}  & s_1^{(c_1,c_2)}  & \cdots & s_{S-1}^{(c_1,c_2)}  \\
    \vdots & \vdots &  \ddots & \vdots \\
    s_S^{(c_1,c_2)}  & s_{S-1}^{(c_1,c_2)}  & \cdots & s_1^{(c_1,c_2)}
\end{bmatrix}
\end{equation}
The structuring is performed by setting all elements to the average of their
corresponding block-subdiagonal and subsequently applying shrinkage
regularization.
This method is simple yet effective, reaching state-of-the-art decoding performance.

### 2024-11-15 chapters/bttda.tex
Proofread mathematical notation

\chapter{Block-term tensor discriminant analysis}
\label{sec:bttda}
\input{chapters/bttda/tensorviz.tex}
\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}

@@

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \hat{\ten{X}}(n)\right)^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

Alternatively, the enhanced performance could also stem from the modeled data
covariance.
Since HODA estimates one within-class scatter matrix
$S_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
model of the data scatter is determined by these per-mode scatter matrices as a
Kronecker product $S_{-1,\text{w}}\otimes S_{-2,\text{w}}\otimes\cdots\otimes S_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
drawn from a multilinear normal distribution~\cite{Ohlson2013}.
However, it is known that the EEG covariance cannot fully be expressed as a
single Kronecker product, but rather is more accurately modeled by a sum of
multiple Kronecker products~\cite{Bijma2005, Sosulski2022}.
Since BTTDA iteratively fits HODA models to the residual error, it will be able
to express the full covariance structure given sufficient blocks.

>>>>

\chapter{Block-term tensor discriminant analysis}
\label{sec:bttda}
\input{chapters/bttda/tensorviz.tex}
\epigraph{%
  ``I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
  Lacking this sum, I therefore borrowed fifty coins from each of my two good friends.
  With the hundred coins, I paid for the beast and received three coins in change.
  Of these, I returned one coin each to my companions, keeping one for myself.
  Now, I owe each friend forty-and-nine coins, and with mine one remaining, the
  total merely equals ninety-and-nine coins, short of one hundred!
  Tell me then, kind stranger, whither hath the missing coin gone?''
}{%
  Ancient Algerian riddle
}
\noindent\emph{This chapter was submitted as~\textcite{VanDenKerchove2024a}.}

@@

A forward model should ensure that reconstruction error is minimized.
In other words, variation captured in the latent tensor should be maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors:
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\text{E}\left[
      \text{vec}\left(\ten{\mathbfcal{E}}(n)\right)\text{vec}\left(\ten{G}(n)\right)
    \right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\}
  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \hat{\ten{X}}(n)\right\]^2
  = \argmin_{\{\mat{A}\}}\sum_n^N\left\[\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right\]^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares algorithm~\cite{Bentbib2022}, iteratively fixing all but one of the activation patterns such that:
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed number of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in \cref{alg:HODA-fw}.
\begin{algorithm}
  \caption[A \acs{hoda} forward solution.]{The \acs{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}

@@

Alternatively, the enhanced performance could also stem from the modeled data
covariance.
Since HODA estimates one within-class scatter matrix
$\mat{S}_{-k,\text{w}}\in\mathbb{R}^{D_k\times D_k}$ per mode, its overall
model of the data scatter is determined by these per-mode scatter matrices as a
Kronecker product $\mat{S}_{-1,\text{w}}\otimes
\mat{S}_{-2,\text{w}}\otimes\cdots\otimes \mat{S}_{-K,\text{w}}$, which corresponds to the assumption that the EEG data is
drawn from a multilinear normal distribution~\cite{Ohlson2013}.
However, it is known that the EEG covariance cannot fully be expressed as a
single Kronecker product, but rather is more accurately modeled by a sum of
multiple Kronecker products~\cite{Bijma2005, Sosulski2022}.
Since BTTDA iteratively fits HODA models to the residual error, it will be able
to express the full covariance structure given sufficient blocks.

### 2024-11-15 chapters/conclusion.tex
Proofread mathematical notation

\chapter{Conclusion \& recommendations}
\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}

>>>>

\chapter{Conclusion \& recommendations}
\epigraph{``The mystery of life isn't a problem to solve, but a reality to
experience.''}{S{\o}ren Kierkegaard}
\noindent\emph{The first paragraph of \cref{sec:conclusion/contrib/decoders} was published as
part of \textcite{VanDenKerchove2022}.}
